---
title: "Lab-02: Pronouns and Data"
author: "Arvind Venkatadri"
date: 06/July/2021
output:
  html_document:
    theme: flatly
    toc: TRUE
    toc_float: TRUE
    toc_depth: 2
    number_sections: TRUE
    code_folding: hide
    code_download: TRUE

abstract: Part of the `R for Artists and Designers` workshop course at the School of Foundation Studies, Srishti Manipal Institute of Art, Design, and Technology, Bangalore.
---


# Introduction

This RMarkdown document is part of the Generic Skills Component  (GSK) of the Course of the  Foundation Studies Programme at Srishti Manipal Institute of Art, Design, and Technology, Bangalore India. The material is based on *A Layered Grammar of Graphics* by Hadley Wickham. The course is meant for First Year students pursuing a Degree in Art and Design. 

The intent of this GSK part is to build Skill in coding in R, and also appreciate R as a way to metaphorically visualize information of various kinds, using predominantly geometric figures and structures.

All RMarkdown files combine code, text, web-images, and figures developed using code. Everything is text; code chunks are enclosed in **fences** (```)


# Goals

- Understand different kinds of data variables
- Appreciate how they can be identified based on the *Interrogative Pronouns* they answer to
- Understand how each kind of variable lends itself to a specific geometric aspect in the data visualization.
- Understand how ask Questions of Data to develop Visualizations


# Pedagogical Note

The method followed will be based on
[PRIMM](https://blogs.kcl.ac.uk/cser/2017/09/01/primm-a-structured-approach-to-teaching-programming/):

-   **PREDICT** Inspect the code and guess at what the code might do,
    **write predictions**
-   **RUN** the code provided and check what happens
-   **INFER** what the `parameters` of the code do and **write comments to explain**. What bells and whistles can you see?
-   **MODIFY** the `parameters` code provided to understand the
    `options` available. **Write comments** to show what you have aimed for and achieved.
-   **MAKE** : take an idea/concept of your own, and graph it.


## Set Up

The `setup` code **chunk** below brings into our coding session **R packages** that provide specific computational abilities and also **datasets** which we can use. 

To reiterate: Packages and datasets are **not** the same thing !! Packages are (small) collections of programs. Datasets are just....information.

# Packages needed

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
library(RColorBrewer)
```

# Introduction

In this RMarkdown document, we try to connect story-making **questions** with two ideas:

a) a *Variable* in a dataset
b) A computed *Quantity* or a *Visual*, based on one or more Variables

So: a question identifies a variable and a question also leads to a data visualization. The idea is to get the intuition behind data, and develop the right questions and hypotheses, that lead to the appropriate direction in exploratory data analysis (EDA) using graphs and charts in R.

<div class="panel panel-success">
  <div class="panel-heading">When it is YOUR TURN:</div>
  <div class="panel-body">
In the following, wherever you see **YOUR TURN**, please respond with explanations, more questions and if you are already confident, code chunks to create new graphs. This will be one of your submissions for this module, on Teams!
  </div>
</div>
 

# Interrogative Pronouns for Data Variables

So how do we ask questions? These are usually with *interrogative pronouns* in English: What? Who? Where? Which? What Kind? How? How Much? How Many? and and so on.

## The `penguins` dataset


```{r Look at Penguins}
names(penguins) # Column, i.e. Variable names
head(penguins) # first six rows
tail(penguins) # Last six rows
dim(penguins) # Size of dataset

# Check for missing data
any(is.na(penguins) == TRUE)
```

1. What are the variable `names()`?
2. What would be the Question you might have asked to obtain that variable?
3. How would you process that variable? 
4. What meta questions could you ask of each variable?
5. Where might the answers take your story?

### YOUR TURN-1




## The `mpg` dataset

```{r Look at mpg}
names(mpg) # Column, i.e. Variable names
head(mpg) # first six rows
tail(mpg) # Last six rows
dim(mpg) # Size of dataset

# Check for missing data
any(is.na(mpg) == TRUE)
```

Look carefully at the variables here. How would you interpret say the `cyl` variable? Is it a number and therefore Quantitative, or could it be something else?

### YOUR TURN-2




## Pronouns and Variables

In the Table below, we have a rough mapping of interrogative pronouns to the kinds of variables in the data:

Pronoun | Answer| English Character | Kind of Data Variable | R Character
--------| ------|------------------ |-----------------------|------------
What  | A Thing | Subject | Nominal| Categorical
Who   | A Person| Subject | Nominal | Categorical
Where | A Place | Subject/Object    | Nominal | Categorical
Whom  | To A Person   | Object| Nominal | Categorical
Which | A Selection or a Type   | Adjective| Qualitative| Qualitative
How   | A Manner/Method| Adverb           | Qualitative | Qualitative
What Kind (1)| Type or Attribute, from a list     | Number (noun/adj) | Qualitative| Factor
What Kind(2) | Type or Attribute, from a list, with order | Affirmative - Comparative - Superlative...| Qualitative with Scale | Factor with Scale
What Kind(3) | Qualitative with Scale, from a list, with order and an absolute value | - | Ordinal | Factor with Scale with "absolute zero"
How Many| A number| Number (noun/adj) | Quantitative| Quantitative
How Often | A number| Number (noun/adj) | Quantitative| Quantitative

Think of :
 - How long? / Since When? : Time
 - How often? / How seldom? : Frequency
 - ..? 

Do take a look at this reference:

[WHAT IS THE DIFFERENCE BETWEEN CATEGORICAL, ORDINAL AND INTERVAL VARIABLES?](https://stats.idre.ucla.edu/other/mult-pkg/whatstat/what-is-the-difference-between-categorical-ordinal-and-interval-variables/)


# Interrogations and Graphs

We can also respond to ( more complex ) Questions, with not just a *variable* but a *data visualization*. This visualization can even involve more than one variable, as we will see

What sort of measures, and visuals charts can we create with different kinds of variables, taken singly or together? Let us write some simple English descriptions of measures and visuals and see what commands they use in R.

Here we will use the Grammar of a package called `ggplot`, which we will encounter in Lab:04. Let us go with our intuition with the code in the following sections. 

Note: since we saw a couple of missing entries in the `penguins` dataset, let us remove them for now. 

```{r}
penguins <- penguins %>% drop_na()
```


## Single Quantitative Variable

1) Questions: How many? How few? How often? How much?

2) Measures: max / min / mean / mode / (units)
  - `max()`, `min()`, `range()`, `mean()`, `mode()`
  
3) Charts: Bar Chart / Histogram / Density  
    - `geom_histogram()` / `geom_density()` 

```{r Single Quant-1}
max(penguins$bill_length_mm)

range(penguins$bill_length_mm, na.rm =TRUE) 
# Note this. Try deleting this option!

ggplot(penguins) + geom_histogram(aes(x = bill_length_mm))
```

```{r Single Quant-2}

ggplot(penguins) + geom_density(aes(bill_length_mm))

```


### YOUR TURN-3






## Two Variables: Quantitative vs Quantitative
We can easily extend our intuition about one quantitative variable, to a pair of them. What Questions can we ask?

1) Questions: How many of this vs How many of that? Does this depend upon that? How are they related? (Remember $y = mx + c$ and friends?)

2) Measures: Correlation / Covariance / T-test / Chi-Square Test for Two Means etc. We won't go into this here!

3) Charts: Scatter Plot / Line Plot / Regression i.e. best fit lines 

```{r Quant vs Quant-1}

ggplot(penguins) + geom_point(aes(x = flipper_length_mm, 
                                  y = body_mass_g))

```


```{r Quant vs Quant-2}
ggplot(penguins) + 
  geom_point(aes(y = body_mass_g , x = flipper_length_mm)) 
# Why do we get a warning?


ggplot(penguins) + 
  geom_point(aes(x = flipper_length_mm, y = bill_length_mm))
```

### YOUR TURN-4






## Single Qualitative/Categorical Variable

1) Questions: Which? What Kind? How?

2) Measures: No of `levels` / Counts for each `level`
  - `counts` / ` tally` (very similar syntax to ggformula)
  
3) Charts: Bar Chart / Pie Chart / Tree Map 
  - `geom_bar` / `geom_bar + coord_polar()` / Find out!!

```{r Single Cat Var-1}
ggplot(penguins) + geom_bar(aes(x = island))
ggplot(penguins) + geom_bar(aes(x = sex))
```


### YOUR TURN-5







## Two Variables: Categorical vs Categorical

What sort of question could we ask that involves *two* categorical variables?

1) Questions: How Many of this Kind(x) are How Many of that Kind(y) ? 

2) Measures: Counts and Tallies sliced by Category
    - `counts` , `tally`
    
3) Charts: Stacked Bar Charts / Grouped Bar Charts / Segmented Bar Chart / Mosaic Chart
    - `geom_bar()` 
    - Use the second Categorical variables to modify `fill`, `color`. 
    - Also try to vary the parameter `position` of the bars. 

```{r Cat vs Cat-1}
ggplot(penguins) + geom_bar(aes(x = island, 
                                fill = species),
                            position = "stack")
```

Storyline: तीन पेनगीन। और तुम भी  तीन

(Oh never mind!)


### YOUR TURN-6






## Two Variables: Quantitative vs Qualitative

Finally, what if we want to look at Quant variables and Qual variables together? What questions could we ask?

1) Questions: How much of this is Which Kind of that? How many vs Which? How many vs How?

2) Measures: Counts, Means, Ranges etc., grouped by Categorical variable. 

3) Charts: Bar Chart using group / density plots by group / violin plots by group / box plots by group
 - `geom_bar` / `geom_density` / `geom_violin` / `geom_boxplot` using Categorical Variable for grouping

```{r Quant vs Qual-1}
ggplot(penguins) + 
    geom_density(aes(x = body_mass_g, 
                 color = island, 
                 fill = island), 
                 alpha = 0.3)
```


```{r Quant vs Qual-2}
ggplot(penguins) + 
  geom_histogram(aes(x = flipper_length_mm,
                 fill = sex))

```



### YOUR TURN-7





# References

1. Data Visualization with R, [Robert Kabacoff](https://rkabacoff.github.io/datavis/)
- Good crisp descriptions of many kinds of graphs, no nonsense book. Available free on the web. 

2. Wickham and Grolemund, [R for Data Science](https://r4ds.had.co.nz/)
 - R Bible. Available free on the web.
 
3. [The best stats you've ever seen | Hans Rosling](https://www.youtube.com/watch?v=hVimVzgtD6w)

 
# Assignments

1. Use the PRIMM process to add comments, questions, inferences to this document, wherever you see YOUR TURN.

2. Create a fresh RMarkdown and similarly analyse one ( or both! ) of the following data sets

- `diamonds` . This dataset is part of the tidyverse package so just type `diamonds` in your code and there it is. 

- `gapminder` !! Yes!!

You will need to install the `gapminder` package to access this dataset.

Ask me for help any time. 


